Bayesian estimation of chronic disease epidemiology from incomplete data

### 2021-10-02 metahit/paper_analyses.Rmd
Fixed hyperparameters / empirical Bayes rework

## Show whether gender effect is common between areas 
## todo pick diseases to illustrate.  IHD, LC biggesc contrast 
## how to show for all ages.

dat <- resall %>%
  filter(var=="cf",
         model %in% c("Hierarchical","Hierarchical joint gender"),
         disease %in% c("Ischemic heart disease", "Stroke", "Lung cancer", "Colon and rectum cancer", "Dementia", "COPD")) %>%
  filter(disease %in% c("Ischemic heart disease", "Lung cancer")) %>%
  mutate(model = fct_recode(model, 
                            "Different between areas" = "Hierarchical",
                            "Common effect" = "Hierarchical joint gender"))
example_age <- 70

#DEBUGME SHOW RAW MORT DATA FOR LC FOR AREAS, SEE IF GENDER DIFFERENCE
gbd %>% filter(ageyr==70, 
               disease=="Tracheal, bronchus, and lung cancer", 
               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)
## YES: MORT/PREV AGREES WITH CF

i <- 3
resi <- readRDS(sprintf("results_hier_gender/res%s.rds",i))
resi <- readRDS(sprintf("results_hier/res%s.rds",3))
resi <- readRDS(sprintf("results_hier/res%s.rds",10))

resi[[1]] %>% 
  filter(var=="cf", between(age, 68, 72), area==1)
resi[[1]] %>% 
  filter(var=="cf", between(age, 68, 72), area==1)

gbd %>% filter(ageyr==70, 
               disease=="Tracheal, bronchus, and lung cancer", 
               area %in% c("Bristol","Liverpool","West Midlands (city region)")) %>%
  group_by(area) %>%
  mutate(mort_num=sum(mort_num), mort_denom=sum(mort_denom), 
         prev_num=sum(prev_num), prev_denom=sum(prev_denom)) %>%
  mutate(p_mort = mort_num/mort_denom, p_prev=prev_num/prev_denom, mpratio=p_mort/p_prev) %>%
  select(sex, area, mort_num, mort_denom, p_mort, p_prev, mpratio)

pdf("../write/gender.pdf", width=6, height=4) # NOTE one for IHD, one for lung cancer

dat %>% filter(age==example_age) %>%
ggplot(aes(x=`50%`, y=area, lty=model, col=gender)) +
  #    coord_cartesian(xlim=c(0.001, 0.2)) +
  facet_wrap(~disease, nrow=1, scales="free_x") + 
  geom_point(position=ggstance::position_dodgev(height=0.6), size=1.5) +
  geom_errorbarh(aes(xmin=`2.5%`, xmax=`97.5%`),
                 position=ggstance::position_dodgev(height=0.6),
                 lwd = 0.7) +
  xlab(sprintf("Case fatality for age %s", example_age) ) + ylab("") + 
  scale_linetype_manual(values = c("Common effect" = "solid",
                                   "Different between areas" = "dashed"
                                   )) +
  labs(col="", lty="Gender effect") +
  theme_bw() + 
  theme(legend.text = element_text(size=7)) +  
  guides(linetype = guide_legend(reverse = TRUE))

@@

# elpd : high is good. positive elpd diff favours additive gender model 
loohg <- looall %>% 
  filter(model %in% c("Hierarchical","Hierarchical joint gender")) %>%
  select(var, area, age, gender, disease, model, elpd_loo) %>%
  pivot_wider(names_from=model, values_from=elpd_loo) %>%
  mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(var, area, gender, disease, agegroup) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

@@

# Time trends

Damn, Absolute incidence estimates are way off both the model results, 
but closest to the non trend one where they match up to age 70.

why.  We are adjusting them to be consistent with prevalence and mortality data. 
But we want to give more weight somehow to current incidence data.  Why tho, if we also believe the current 
prev and mort data? are we trying to make them all fit into a bad model?

Right the fitted values for year 2010 and earlier match up with current data.  There's a big jump around 2010

But then they will carry on changing through time in the future if we use them in a model

pdf("../write/trend_res.pdf")

datt <- gbd %>% 
  filter(area=="Leeds", disease=="Ischemic heart disease") %>%
  mutate(inc_prob = inc_num / inc_denom, 
         inc_rate = -log(1-inc_prob), 
         var = "Incidence", 
         gender = sex)

resoy <- resall %>% filter(model=="Time trends", 
                            year == 2017,
                            area=="Leeds", 
                            disease=="Ischemic heart disease",
                            var %in% c("inc")) %>% mutate(var="Incidence")

rest <- resall %>% filter((model == "Independent areas") | (model=="Time trends" & year==2017),
                          area=="Leeds", 
                          disease=="Ischemic heart disease",
                          var %in% c("inc","cf")
#                          gender=="Male"
) %>% full_join(resoy)

resopt <- readRDS(file="resopt_trend.rds") %>% 
  filter(disease=="Ischemic heart disease",area=="Leeds",
         var %in% c("inc","cf"), year==99) %>%
  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"))

rest %>% 
  filter(between(age, 50, 90)) %>% 
  mutate(var = fct_recode(var, "Incidence"="inc", "Case fatality"="cf"),
         model = fct_recode(model, "No time trends" = "Independent areas")) %>%
ggplot(aes(x=age, y=`50%`, col=model, fill=model, 
           group=interaction(gender, model))) +
  coord_cartesian(ylim=c(0.0,0.04)) +
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.2) +
  facet_wrap(vars(gender, var), nrow=2, scales="free_y") +
  geom_line() + 
  xlab("Age") + ylab("") + 
  labs(col="", fill="") + 
  geom_line(data=datt, aes(x=ageyr, y=inc_rate), inherit.aes = FALSE) +
  geom_line(data=resopt, aes(x=age, y=par), inherit.aes=FALSE, col="blue")

## Point estimates are accurate, so we can use them to diagnose the bias

#  geom_line(data=resoy, aes(x=age, y=`50%`, col=factor(year), group=year), lty=2, inherit.aes = FALSE)
dev.off()

## Experimental sensitivity analyses to different trends
## what was it? 2003-2010 continues. saved in restrs 
## where's the conclusion from this?  estimates vary a lot

rest_saved <- rest %>% 
  filter(between(age, 50, 90), model=="Time trends", 
         gender=="Male", var %in% c("inc","cf")) 
res_saved <- resnh %>% filter(between(age, 50, 90), model=="Independent areas", disease=="Ischemic heart disease", area=="Leeds", gender=="Male", var %in% c("inc","cf"))

restrs <- restrs %>% 
  mutate(trendin = ifelse(inctrend & cftrend, "both", 
  ifelse(inctrend, "inc", ifelse(cftrend, "cf", "neither")))) %>%
  mutate(trendin = ifelse(trsens=="zero", "neither", trendin))

dat <- restrs %>% 
  filter(var %in% c("inc","cf"), model=="smoothed", trsens %in% c("hi", "zero"))
dat$age <- rep(1:100, length.out=nrow(dat))
dat %>%
  filter(age < 80) %>%
ggplot(aes(x=age, y=`50%`, col=trendin)) + 
  geom_line(lwd=2) + 
  facet_grid(cols=vars(var)) + 
  geom_line(data=datt %>% filter(gender=="Male") %>% mutate(var=fct_recode(var,"inc"="Incidence")), 
            aes(x=ageyr, y=inc_rate), inherit.aes = FALSE)

+ 
  geom_line(data=rest_saved, col="black") + 
  geom_line(data=res_saved, col="blue")

restrs %>% 
  filter(var %in% c("inc","cf")) %>%
  mutate(age = rep(1:100, length.out=n())) %>%
  filter(model=="smoothed", age==60, var=="cf", cftrend==TRUE, inctrend==TRUE)

Right this is saying that it doesn't make much difference to CF, but big diff to inc as in Scar
were the original plots wrong.   OK fine just axis limit distortion.

So the only problem left is disagreement with current empirical incidence 
dismod2 types would say the model is "forcing consistency" 
evidence of conflict between current inc and current CF  
Hm is this evidence that the gbd data are consistent with themselves under a model that ignores trends.

Hey wot it does match.   Wot about original independent-areas results

cf - higher with inc or both, lower with cf or neither 
inc - higher with cf only, then neither, then both, then lowest with inc.

So dose response is OK 
but they interact in unpredictable ways 
and need to fix the 5 year tihng 
and reconcile the modest changes here with the big changes from the real data

Compare real with artificial trend data.  Where does the difference come from? 
Looks reasonably representative of CF to me.  Age 70 value is generous 
Incidence: trend data is more noisy.  Pick an even higher value that's on the high end of the data.

```{r}
ism <- inctrends_hi %>% as.data.frame() %>% 
  mutate(age = 1:100) %>%
  pivot_longer(names_to = "year", values_to = "p2017", cols = 1:100) %>%
  mutate(year = as.numeric(year)) %>% 
  filter(age %in% c(40,60,80,100), year > 1970) %>% 
  mutate(age = factor(age), model = "sens")

ihdtrends %>% 
  filter(age %in% c(40,60,80,100), year > 1970, gender=="Male", outcome=="Incidence") %>% 
  mutate(age = factor(age), model = "data") %>%
  full_join(ism) %>%
  ggplot(aes(x=year, y=p2017, col=age, lty=model)) + geom_line()

Well that's odd.

With inc trends on. 
There's much less of an effect than there was with the real data. 
Consistent dose response effect 
CF trend makes no difference to current CF ests , but affects current inc ests. Wot?

With cf trends on
Only an effect on cf if there is inc trends.
Consistent dose response when there's an effect
Inc est: both inc and cf trends make a difference.  moderate differences

Is this a bug in the sens analysis setup??   cf trends are not affecting cf
And why are there still those stupid 5 year artefacts for incidence?  didn't i transfer the new data?

OK think og it a new way 
TREND IN: 
inc but not cf : 
cf but not inc : reduced inc 
both inc and cf:     reduced inc

inc highest: 
inctrend false, cftrend true
inctrend true, cftrend true 
cftrend false.   
inc lowest

cf highest with 
inctrend true, cftrend false or true 
inctrend false

>>>>

## Get curves of m/f CF risk ratios by age and area, from the hier model with 
## different gender effect for each area 
reshi <- resh %>%  
  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
         area %in% cityregions, 
         var=="cf") %>%
  mutate(est = log(`50%`), se=(log(`97.5%`)-log(`2.5%`))/4) 
reshie <- reshi %>%
  select(age, gender, area, disease, est) %>% 
  pivot_wider(names_from=gender, values_from=est) %>%
  mutate(lhrmale = Male - Female) %>%
  select(age, area, disease, lhrmale)
reshis <- reshi %>%
  select(age, gender, area, disease, se) %>% 
  pivot_wider(names_from=gender, values_from=se) %>%
  mutate(sediff = sqrt(Male^2 + Female^2)) %>%
  select(age, area, disease, sediff) %>%
  left_join(reshie, by=c("age","area","disease")) %>%
  mutate(est = exp(lhrmale), 
         lower=exp(lhrmale - 2*sediff),
         upper=exp(lhrmale + 2*sediff))

## Superimpose the equivalent common curve from the common effect model
resgi <- resg %>%  
  filter(! disease %in% c("Breast cancer", "Stomach cancer", "Uterine cancer"), 
         area %in% cityregions, 
         var=="cf") %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
  mutate(est = log(`50%`), se=(log(`97.5%`)-log(`2.5%`))/4) 
resgie <- resgi %>%
  select(age, gender, area, disease, est) %>% 
  pivot_wider(names_from=gender, values_from=est) %>%
  mutate(lhrmale = Male - Female) %>%
  select(age, area, disease, lhrmale)
resgis <- resgi %>%
  select(age, gender, area, disease, se) %>% 
  pivot_wider(names_from=gender, values_from=se) %>%
  mutate(sediff = sqrt(Male^2 + Female^2)) %>%
  select(age, area, disease, sediff) %>%
  left_join(resgie, by=c("age","area","disease")) %>%
  mutate(est = exp(lhrmale), 
         lower=exp(lhrmale - 2*sediff),
         upper=exp(lhrmale + 2*sediff))

## Put this in the paper 
## Hazard ratio for men compared to women for case fatality 
## This is good evidence that the general pattern is common among areas 
## TODO Confirm this or otherwise by LOO CV

pdf("../../write/gender_hr.pdf")
ggplot(reshis %>% filter(age > 60), 
       aes(x=age, y=est, group=area, col=area)) + 
  geom_ribbon(aes(ymin=lower, ymax=upper), fill="gray", col=NA, alpha=0.2) + 
  geom_hline(yintercept = 1) +
  geom_line(lwd=0.8) +
  geom_line(data=resgis %>% filter(age>60), col="black", lty=2) +
  facet_wrap(~disease) +
  labs(col="Area") +
  coord_cartesian(ylim=c(0,3)) + 
  ylab("Case fatality rate ratio (men/women)") + xlab("Age (years)")
dev.off()

@@

## FIXME rerun hier with bugfix in saving loo output .  was triggered when we changed rundf
## to exclude (stmc and?) utrc

looall %>% filter(var=="mort", area=="Bristol", gender=="Male", disease=="Lung cancer", 
                  model=="Hierarchical", age==1)

# elpd : high is good. positive elpd diff favours additive gender model 
loohg <- looall %>% 
  filter(model %in% c("Hierarchical","Hierarchical joint gender")) %>%
  select(var, area, age, gender, disease, model, elpd_loo) %>%
  pivot_wider(names_from=model, values_from=elpd_loo, values_fn=length)

%>%
  mutate(elpd_diff = `Hierarchical joint gender` - `Hierarchical`) %>%
  mutate(agegroup = cut(age, seq(0,100,5), right=FALSE, include.lowest = TRUE)) %>%
  group_by(var, area, gender, disease, agegroup) %>%
  summarise(elpd_diff = sum(elpd_diff), .groups=c("drop")) %>%
  mutate(agemid = as.numeric(as.character(factor(agegroup, labels = seq(2.5, 97.5, by=5)))))

@@
